---
title: "Run Expectancy (Chapter 5)"
output: pdf_document
---
```{r, echo=FALSE}
setwd("~/Documents/baseballStats/runExpectancy")
```
# Introduction


# Prepare the Data

## Runs Scored
The data for the 2011 season is housed in the \texttt{all2011.csv} file in the \texttt{data} folder. The field names are stored in a file called \texttt{fields.csv} in the same folder. Let's load them in and set the header with the \texttt{fields} file.
```{r, cache=TRUE}
data2011 <- read.csv("../data/all2011.csv", header=FALSE)
fields <- read.csv("../data/fields.csv")
names(data2011) <- fields[, "Header"]
```

The loaded dataset has `r dim(data2011)[2]` attributes and `r dim(data2011)[1]` observations.

What we're after is the average number of runs scored in the rest of the inning with each given combination of outs and runners. That's 3 outs times $2^3$ base combinations, for 24 base/out states. We want to make a new variable 
\[
\text{RUNS.ROI} = \text{Total Runs Scored in Inning} - \text{Current Runs Scored}
\]

And average that variable across each base/out state.

```{r}
data2011$RUNS <- data2011$AWAY_SCORE_CT + data2011$HOME_SCORE_CT
```

Defines the current total runs scored at each plate appearance. We'll want to uniquely identify each half inning using a new variable \texttt{HALF.INNING}

```{r}
data2011$HALF.INNING <- paste(data2011$GAME_ID, data2011$INN_CT, data2011$BAT_HOME_ID)
```

Any time a runner or batters destination id is greater than 3, a run scores, these are described in the variables \texttt{BAT\_DEST\_ID} and \texttt{RUN1\_DEST\_ID}, with \texttt{RUN1} being the runner on first, \texttt{RUN2} being on second, etc.

```{r}

data2011$RUNS.SCORED <- (data2011$BAT_DEST_ID > 3) + 
                        (data2011$RUN1_DEST_ID > 3) + 
                        (data2011$RUN2_DEST_ID > 3) + 
                        (data2011$RUN3_DEST_ID > 3)
```

This is next part is incredibly slick. We run two aggregations, one aggregating total runs scored across the innings, i.e. summing all the values of \texttt{RUNS.SCORED} across plate appearance, grouped by \texttt{HALF.INNING}. Next we do an aggregation using the "[" function, which is the subset function, passing it a value of 1 for its first argument. In other words, evaluate `x[1, ]` for each `x`, grouped by \texttt{HALF.INNING}.
```{r}
RUNS.SCORED.INNING <- aggregate(data2011$RUNS.SCORED, list(HALF.INNING=data2011$HALF.INNING), sum)
RUNS.SCORED.START <- aggregate(data2011$RUNS, list(HALF.INNING=data2011$HALF.INNING), "[", 1)
```

Now that we have these two vectors, we can calculate the max runs in a given half inning 

```{r}
MAX <- data.frame(HALF.INNING=RUNS.SCORED.START$HALF.INNING)
MAX$x <- RUNS.SCORED.INNING$x + RUNS.SCORED.START$x
data2011 <- merge(data2011, MAX)
N <- ncol(data2011)
names(data2011)[N] <- "MAX.RUNS"

data2011$RUNS.ROI <- data2011$MAX.RUNS - data2011$RUNS
```

And there you have it, total runs in the inning minus current runs, just as we defined it in the beginning.

## Base/Out State
To deterimine the base/out states, we first determine the bases
```{r}
RUNNER1 <- ifelse(as.character(data2011[, "BASE1_RUN_ID"]) == "", 0, 1)
RUNNER2 <- ifelse(as.character(data2011[, "BASE2_RUN_ID"]) == "", 0, 1)
RUNNER3 <- ifelse(as.character(data2011[, "BASE3_RUN_ID"]) == "", 0, 1)
```

The above vectors are the same length as our dataframe, with a 0 if the respective base was empty for the plate appearance, and 1 if there was a runner on. We then write a function to combine those three vectors with the outs

```{r}
get.state <- function(runner1, runner2, runner3, outs){
  runners <- paste(runner1, runner2, runner3, sep="")
  paste(runners, outs)
}

data2011$STATE <- get.state(RUNNER1, RUNNER2, RUNNER3, data2011$OUTS_CT)
```

We only care about plays were there is a change in state, so using the variables available to us we'll calculate the state after the play.

```{r}
NRUNNER1 <- as.numeric(data2011$RUN1_DEST_ID == 1 | 
                         data2011$BAT_DEST_ID == 1)

NRUNNER2 <- as.numeric(data2011$RUN1_DEST_ID == 2 | 
                         data2011$RUN2_DEST_ID == 2 | 
                         data2011$BAT_DEST_ID == 2)

NRUNNER3 <- as.numeric(data2011$RUN1_DEST_ID == 3 | 
                         data2011$RUN2_DEST_ID == 3 | 
                         data2011$RUN3_DEST_ID == 3 | 
                         data2011$BAT_DEST_ID == 3)

NOUTS <- data2011$OUTS_CT + data2011$EVENT_OUTS_CT

data2011$NEW.STATE <- get.state(NRUNNER1, NRUNNER2, NRUNNER3, NOUTS)

```

We throw away the plays where there was no change in runs or state -- I looked into it, they're mostly errors on foul flies, and errors on picked off runners, where they end up back at the same base.

```{r}
data2011 <- subset(data2011, (STATE != NEW.STATE) | (RUNS.SCORED > 0))
```

Last thing to take care of is walkoffs, where a run is scored but the game ends without getting 3 outs. We can take care of those with \texttt{ddply} in the \texttt{plyr} library. We create a new dataframe called \texttt{data.outs} which is the total outs in each \texttt{HALF.INNING}. Then, we merge that back with our main data frame and filter off any that are less than 3.

```{r, cache=TRUE}
library(plyr)

data.outs <- ddply(data2011, .(HALF.INNING), summarize, Outs.Inning=sum(EVENT_OUTS_CT))
data2011 <- merge(data2011, data.outs)
data2011C <- subset(data2011, Outs.Inning == 3)

```


## Expected Runs

At last we can compute the expected runs for each base/out state. Note we parse the grouping variable (state) to get the number of outs and order by outs.

```{r}
RUNS <- aggregate(data2011C$RUNS.ROI, list(data2011C$STATE), mean)
RUNS$Outs <- substr(RUNS$Group, 5, 5)
RUNS <- RUNS[order(RUNS$Outs), ]
```

We can now make a matrix 

```{r}
RUNS.out <- matrix(round(RUNS$x, 2), 8, 3)
dimnames(RUNS.out)[[2]] <- c("0 outs", "1 out", "2 outs")
dimnames(RUNS.out)[[1]] <- c("000", "001", "010", "011", "100", "101", "110", "111")

```

# The Matrix
The table is given in Table \ref{REMatrix}

```{r echo=FALSE,results='asis',warning=FALSE}
library(xtable)

options(xtable.comment = FALSE)
xtable(RUNS.out, caption="Run Expectancy Matrix for 2011", label="REMatrix")


```